---
title: "Assignment 2. Part a. SQL and R - Movie ratings analysis."
author: "Maxim Arisov"
format: html
date: 09/12/26
code-fold: true
toc: true
editor: visual
---

# SQL and R - Movie Ratings Analysis

# Introduction. Approach

## Creation of SQL Database using the PostgreSQL

For this assignment, I collected movie ratings from different users. The ratings are from 1 to 5. I used three tables: users, movies, and ratings. The ratings table connects each user with the movies they rated. For this particular assignment 2, part a, I will create the SQL Database based on the provided instructions (using CREATE TABLE, etc.). I will show the creation of data in the Excel sheets, first, and then transferring it to CSV files, and eventually uploading to the PostgreSQL.

## Creation of CSV file from a SQL Query. R use.

I will then create another CSV files from PostgreSQL Query and will use it in R studio to count ratings per user and movie, compute average ratings, identify and clean any missing values, and calculating means and counts. Dplyr will be implemented during coding process. First, I stored the data using SQL. I used separate tables to keep the data organized. Second, I loaded the data into R. I combined the tables using `user_id` and `movie_id`. Then, I used R to check and analyze the data. I calculated the average rating for each movie and the number of ratings for each movie. For the assignment 2a, I also checked for missing ratings. I kept missing ratings as `NA` because a missing value means the user did not give a rating.

# Code Base

This project analyses movie ratings obtained from participants. The data is stored using users, movies, and ratings.

```{r}
movies <- read.csv("https://raw.githubusercontent.com/Maximus2485/Assignment-2.-part-a.-movie-ratings/refs/heads/main/movies.updated.csv")
ratings <- read.csv("https://raw.githubusercontent.com/Maximus2485/Assignment-2.-part-a.-movie-ratings/refs/heads/main/movie%20ratings.updated.csv")
users <- read.csv("https://raw.githubusercontent.com/Maximus2485/Assignment-2.-part-a.-movie-ratings/refs/heads/main/users.%20updated.csv")

head(movies)
head(ratings)
head(users)

```

**Lets install the SQL packages now**

```{r}
library(DBI)
library(RSQLite)

con <- dbConnect(SQLite(), "movie_ratings.db")
```

**Lets connect to SQLite**

```{r}
library(DBI)
library(RSQLite)

con <- dbConnect(RSQLite::SQLite(), "movie_ratings.db")

dbIsValid(con)
```

**Lets create SQL tables in SQLite**

```{sql connection=con}
CREATE TABLE IF NOT EXISTS users1 (
user_id INT PRIMARY KEY,
name TEXT
);

```

```{sql connection=con}
CREATE TABLE IF NOT EXISTS movies1 (
movie_id INT PRIMARY KEY,
title TEXT
);

```

```{sql connection=con}
CREATE TABLE IF NOT EXISTS ratings1 (
user_id INT, movie_id INT, rating INT,
FOREIGN KEY (user_id) REFERENCES users (user_id),
FOREIGN KEY (movie_id) REFERENCES movies(movie_id)
);
```

**Lets run a command to see whether the tables were created**

```{r}
DBI::dbListTables(con)
```

**Lets check now the data variables**

```{r}
DBI::dbListFields(con, "movies1")
DBI::dbListFields(con, "users1")
DBI::dbListFields(con, "ratings1")
```

**Lets populate SQL tables**

```{r}
dbWriteTable(con, "users1", users, overwrite = TRUE)
dbWriteTable(con, "movies1", movies, overwrite = TRUE)
dbWriteTable(con, "ratings1", ratings, overwrite = TRUE)
```

**Lets check missing ratings in SQL**

```{sql connection=con}
SELECT *
FROM ratings1 
WHERE rating is NULL
```

**Lets join the 3 tables with SQL**

```{sql connection=con}
SELECT
    u.user_id,
    u.name,
    m.movie_id,
    m.title,
    r.rating
FROM ratings1 AS r
JOIN users1 AS u
    ON r.user_id = u.user_id
JOIN movies1 AS m
    ON r.movie_id = m.movie_id;


```

**Now, Lets bring the SQL results into the R**

```{r}
movie_data <- DBI::dbGetQuery(con,
                              "SELECT
    u.user_id,
    u.name,
    m.movie_id,
    m.title,
    r.rating
FROM ratings1 AS r
JOIN users1 AS u
    ON r.user_id = u.user_id
JOIN movies1 AS m
    ON r.movie_id = m.movie_id;
"
                              )
head(movie_data)
```

**Lets analyse the data in R**

```{r}
#Lest check the average rating for each movie (mean)
aggregate(
  rating ~ title, 
  data = movie_data,
  FUN = mean,
  na.rm = TRUE
)
```

```{r}
#Lets check and find missing ratings
sum(is.na(movie_data$rating))
```

```{r}
#Lets find number of ratings for each movie
aggregate(
  rating ~ title, 
  data = movie_data,
  FUN = function(x) sum(!is.na(x))
)
```

```{r}
#Lets check number of ratings for user
aggregate(
  rating ~ name,
  data = movie_data,
  FUN = function(x) sum(!is.na(x))
)
```

## Conclusion

During this assignment 2. part a, I stored movie rating data in a SQL database and used SQL subsequently to join the tables. Then, I loaded the obtained data into the R and analyzed the ratings. At the end, I checked for missing values and calculated the average rating, number of ratings for each movie.

### **Use of resources**

The R for Data Science (2e) in addition with Gemini AI were used to find some issues and shortcomings.

GitHub raw data is available at the links below:

<https://raw.githubusercontent.com/Maximus2485/Assignment-2.-part-a.-movie-ratings/refs/heads/main/movie%20ratings.updated.csv>

<https://raw.githubusercontent.com/Maximus2485/Assignment-2.-part-a.-movie-ratings/refs/heads/main/movies.updated.csv>

<https://raw.githubusercontent.com/Maximus2485/Assignment-2.-part-a.-movie-ratings/refs/heads/main/users.%20updated.csv>

## References

Google DeepMind. (2025). Gemini 3 Flash \[Large language model\].\
\[https://gemini.google.com.\](https://gemini.google.com/) Accessed September 12th, 2026.

Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). ***R for data science*** (2nd ed.). O'Reilly Media. https://r4ds.hadley.nz/
